### 2013-11-20 patchy-selection-paper.tex
.

Consider a population spread across a landscaper, 
within which are patches of habitat to which individuals are (initially) poorly adapted,
surrounded by large areas to which the population is well adapted.
(When we refer to ``patches'' it is to these pieces of habitat.)
Suppose it takes only a single mutational change to create an allele ($B$) that adapts an individual to the poor habitat type, 
a change that occurs at a (low) rate of $\mu$ per chromosome per generation. 
We assume that this initially rare new
allele has fitness $1+s_p$ relative to the unmutated type ($b$) in these ``poor'' habitat patches,
and assume that it has fitness $1-s_m$ when rare in the intervening areas, with $s_p, s_m > 0$.
To simplify matters, we assume that the disadvantage $s_m$ 
is sufficiently large that, on the relevant timescale,
the allele is very unlikely to fix locally in the regions where it is at a disadvantage.
(The case where $s_m=0$ requires a different approach, which we do not treat here.)

@@

\paragraph{Previous work} 
We will make use of several previous results from the literature. 
\citet{slatkin1973geneflow}, working in a one dimensional, deterministic framework, 
showed that if the physical width of the patch is less than $2 \tan^{-1} (\sqrt{s_m/s_p})$, 
then there is no stable equilibrium with both $b$ and $B$ present,
so that migrational swamping prevents $B$ from establishing \citep[see also][ for a review]{lenormand2002limits}.
\citet{barton1987establishment}, using general theory of \citet{pollak1966survival}, showed (again in one dimension) 
that this critical patch size also holds in a stochastic framework.
\citet{barton1987establishment} also showed that
for patches above this critical size,
the probability of establishment of a new mutant that appears at distance $r$ from the patch
decays exponentially with distance from the patch, with the scale given by $\sigma/\sqrt{s_m}$,
and that mutations appearing within the patch have probability of establishment
less than the probability for a panimictic population,
but approaching this as the size of the patch increases.
(This latter panmictic probability of establishment we denote $p_e$,
and often approximate by $2 s_p / \xi^2$ \citep{haldane,fisher}.)
This result holds quite generally for a geographically spread population that experiences a uniform selection
pressure \citep{maruyama1970fixation,cherry2003diffusion}.

@@

\begin{figure}[ht!]
  \begin{center}
    \includegraphics{sim-snapshots}
  \end{center}
  \label{fig:sim_snapshots}
  \caption{
  Results from individual-based simulation, on a one-dimensional grid:
  the focal allele is beneficial in the central 20 demes, with $s_b=0.1$,
  and deleterious elsewhere, with $s_m=-0.02$.
  The dynamics were simulated for 10000 generations; demes had 1000 individuals in each;
  dispersal was to nearby demes with a mean dispersal distance of $\sigma \approx 1.66$  times the deme spacing.
  \textbf{(A)} Allele frequencies across space: colored lines are snapshots at an evenly spaced set of times,
  and the solid lines shows the cumulative mean occupation frequency across all generations.
  \textbf{(B)} A segment of the temporal dynamics: space is again shown across the horizontal axis;
  with time on the vertical axis; demes are colored darker red the greater then allele frequency at the corresponding time.
  The 50\% and 5\% frequency contour lines are in grey,
  and the genealogy of 50 individuals in a rare long-distance excursion is traced back with black lines.
  }
\end{figure}

@@

The arguments depends on the following decomposition of the genealogy of $B$ alleles.
First, fix some intermediate distance $r_0$ from the established patch,
and classify all $B$ alleles present at any time outside $r_0$ into families
if they share a common ancestor living outside $r_0$.
If $r_0$ is large enough that there are not too many families
and interactions between family members are weak,
then these families can be approximated by a collection of independent spatial branching processes.
(This can be made formal in the limit of large population density.)
We can extend these families by 
including any individual outside the original patch (but perhaps inside $r_0$) 
that has a parent outside $r_0$,
so that these branching processes are rooted at $r_0$, and are killed upon reaching the original patch.
This decomposition is illustrated in~\ref{fig:sim_snapshots}, and later in figure~\ref{fig:branching_decomp}.
In terms of these families, the equilibrium frequency above is then
\begin{align}
    \label{eqn:gestalt_q}
    \begin{split}
        q(r) &= \text{ ( outflux of families ) } \times \text{ ( mean occupation density of a family at $r$ ) } ,
\end{split}
\end{align}
and the quantity we wish to compute,
the effective rate of at which \emph{families of} migrant $B$ alleles establish in a new patch at distance $r$, 
is
\begin{align}
    \label{eqn:gestalt_migrate}
    \begin{split}
        \migrate(r) &= \text{ ( outflux of families ) } \times \text{ ( probability a family establishes in patch at $r$ ) }
    \end{split}
\end{align}
All terms except the ``outflux of families'' are simple calculations with spatial branching processes;
and we can solve for this outflux using the established form for \eqref{eqn:gestalt_q}.
Here $\migrate$ reflects the mean number of establishing migrant families per unit time,
rather than individuals,
because family establishment events may include more than one individual,
so individual establishment events are clustered in time,
and using their overall density to obtain a typical waiting time until the next event
will underestimate that time.

@@

To understand the process of adaptation by migration,
we therefore need to look more closely at the rare families of $B$ the allele far from the patches where it is established.
Consider one such family.
As noted above, we can model this by assuming that each individual
migrates, then reproduces, independently of all other family members.
The mean squared migration distance is $\sigma^2$, takes the migrant in a random direction,
and if migration takes the individual into a patch where the $B$ allele is favored, 
we drop the individual from our family.
Since the $B$ allele is uniformly deleterious elsewhere,
we can decouple reproduction and migration
by first forming a tree in which each individual has a random number of offspring
(with mean $1-s_m$ and variance $\xi^2$),
then assigning spatial locations to each individual in the tree,
and then pruning any branches that happen to wander into a patch.
This pruning removes both branches leading back into the already established patch (which are irrelevant)
and branches leading into the new patch (we will need to count these).
Since we are concerned with the rare migrant families that transit quickly between patches, 
backmigrants to the already established patch will contribute little, since they are going the wrong direction;
so to simplify the analysis we ignore this detail.  % but could add this detail in an appendix

Since each individual has on average $(1-s_m)$ offspring,
a family has on average $(1-s_m)^t$ members in generation $t$.
Let $p_e(t)$ be the chance of extinction by time $t$, and let $K_t$ be the (random) family size conditioned on nonextinction,
so that $(1-s_m)^t = (1-p_e(t))\E[K_t]$.
It turns out that by assuming the fate of each offspring is independent (i.e.\ that this forms a branching process),
and that the distribution of offspring numbers is not too heavy-tailed \citep[see][for details]{jagers1975branching},
that $K_t$ has a limiting distribution: $K_t \dconv K$ 
and that $(1-s_m)^{t}/(1-p_e(t)) \to \E[K]$ as $t \to \infty$.
In other words, the chance of survival to time $t$ is a constant multiplied by $(1-s_m)^t$,
and conditional on survival, the family size is fixed.

We can understand this quite concretely, by a method described rigorously by \citet{geiger1999elementary}.
When we condition on survival until $t$, we condition on existence of at least one lineage from time $0$ to time $t$.
Once given this lineage, nothing distinguishes any of the other reproduction events --
each individual in this lineage must give birth to at least one offspring,
and all other reproduction events are unconditioned.
In other words, the genealogy of a family that survives until $t$
can be decomposed into a ``trunk'' that lasts until $t$
sprouting independent ``branches'' that may or may not survive until $t$.
This process is depicted in Figure~\ref{fig:branching_decomp}A.
Concretely, if we write $Z_t$ for the number of offspring in the branching process alive at time $t$
(so $\P\{Z_t = 0\} = p_e(t)$),
then
\begin{align}
  \left( Z_t \; \vert \; Z_t>0 \right) \deq \sum_{s=0}^t \sum_{k=1}^{W_s-1} Z^{(s,k)}_{t-s},
\end{align}
where each $Z^{(s,k)}$ is an independent copy of $Z$,
and each $W_s$ is an independent copy of the offspring distribution, conditioned on $W_s \ge 1$.

@@

This suggests that we add in the spatial motion by following only the motion of the ``trunk''
(i.e.\ the red line in figure \ref{fig:branching_decomp}),
whose motion is is approximately Brownian, with variance $\sigma^2$ per generation.
(Recall $\sigma$ is the dispersal distance.)
Suppose that the new patch is at distance $r$.
The probability density for the first time such a Brownian motion $B_s$ hits $r$ is
\citep[XXX]{feller}
\plr{work in $d=2$}
% note $\sigma B_t \deq B_{\sigma^2 t}$ so $\P\{ \sup_{s\le t} \sigma B_s \le r \} = \P\{ \sup_{s \le \sigma^2 t} B_s \le r \}
\begin{align} \label{eqn:hitting_dist}
  \P\{ B_t=r \;\mbox{and} \; B_u<r \;\mbox{for all}\; u<t\} =  \frac{r}{\sigma^3 t^{3/2}\sqrt{2\pi}} \exp\left(-\frac{r^2}{2t\sigma^2}\right) .
\end{align}
Let $\tau$ be this random time at which a given migrant family hits the new patch,
with the convention that $\tau =\infty$ if the family never gets here.
We approximate the chance that a given family arrives at the new patch at time $t$, i.e.\ that $\tau=t$,
by the product of equation~\eqref{eqn:hitting_dist} and the probability that the migrant family survives until time $t$.
For convenience, define $\alpha := - \log(1-s_m)$ to be the rate of decrease between patches (note that $\alpha \approx s_m$),
and $\mu_K = \E[K]$ to be the limiting mean family size conditioned on survival,
so that $1-p_e(t) \simeq (1/\mu_K) \exp(-\alpha t)$.
Then, the probability distribution of $\tau$ is given by
\begin{align}
    \P\{ \tau \in dt \} =  \frac{1}{\mu_K} \frac{r}{\sigma^3 t^{3/2}\sqrt{2\pi}} \exp\left(-\frac{r^2}{2t\sigma^2} - \alpha t \right) dt  .
\end{align}

@@

The probability of establishment of an outmigrant from the already-established patch is 
equal to the probably the trunk arrives (equation~\eqref{eqn:estab_prob}) multiplied by the probability it establishes (equation~\eqref{eqn:prob_estab_arrival}),
namely
\begin{align} \label{eqn:bp_prob_estab}
    F(r) &:= \text{( prob family establishes at $r$ )} = \frac{2s_p}{\xi^2} \exp\left( - \frac{\sqrt{2 \alpha} r}{\sigma} \right) ,
\end{align}

@@

Now, we can obtain the effective ``outflux of families'' from information about the migrant families, combined with \eqref{eqn:gestalt_q}.
Define $\gamma$ to be this outflux; 
i.e.\ the mean number of individuals per unit time that are born carrying the $B$ allele closer than $r_0$ to the original patch
but migrate to a larger distance.
Then for $r>r_0$, the mean occupation density $q(r)$ is equal to $\gamma$ 
multiplied by the mean occupation density of the family descending from such a migrant,
which we denote $u(r)$.
Denote by $Z_t$ the number of individuals such a family has after time $t$,
and the locations of these by $X_1(t), \ldots, X_{Z_t}(t)$.
Then this mean occupation density, at $x$, is the expected number of individuals ever present at $x$. 
Since the expected number of individuals present at time $t$ is $e^{\alpha t}$
and the location of each of these individuals is marginally distributed as a  Gaussian with variance $\sigma^2 t$,
we can write this expectation as
\begin{align}
    u(x) &= \int_0^t \E[ \#\{ k : \; X_k(t) \in dx \} ] dt \\
    &= \int_0^t e^{\alpha t} \frac{1}{\sqrt{2 \pi t \sigma^2}} e^{-\frac{x^2}{2t\sigma^2}} dt \\
    &= \frac{ 1 }{ \sigma \sqrt{2 \alpha} } \exp\left( - \frac{ x \sqrt{2\alpha} }{ \sigma } \right) 
\end{align}
\plr{check that}
The integral is evaluated in Appendix~\ref{apx:some_integrals}.
\plr{Also examine non-dependence on $r_0$.}

>>>>

Consider a population spread across a landscape, 
within which are patches of habitat to which individuals are (initially) poorly adapted,
surrounded by large areas to which the population is well adapted.
(When we refer to ``patches'' it is to these pieces of habitat.)
Suppose it takes only a single mutational change to create an allele ($B$) that adapts an individual to the poor habitat type, 
a change that occurs at a (low) rate of $\mu$ per chromosome per generation. 
We assume that this initially rare new
allele has fitness $1+s_p$ relative to the unmutated type ($b$) in these ``poor'' habitat patches,
and assume that it has fitness $1-s_m$ when rare in the intervening areas, with $s_p, s_m > 0$.
To simplify matters, we assume that the disadvantage $s_m$ 
is sufficiently large that, on the relevant timescale,
the allele is very unlikely to fix locally in the regions where it is at a disadvantage.
(The case where $s_m=0$ requires a different approach, which we do not treat here.)

@@

\paragraph{Previous work} 
We will make use of several previous results from the literature. 
\citet{haldane1948theory}, working with the one dimensional partial differential equation,
showed that with large patches, the spatial scale over which the two types intergrade is given by $\sigma/\sqrt{s}$,
with $s$ either $s_m$ or $s_p$ as appropriate.
\citet{slatkin1973geneflow}, in the same setting,
showed that if the physical width of the patch is less than $2 \tan^{-1} (\sqrt{s_m/s_p})$, 
then there is no stable equilibrium with both $b$ and $B$ present,
so that migrational swamping prevents $B$ from establishing \citep[see also][ for a review]{lenormand2002limits}.
\citet{barton1987establishment}, using general theory of \citet{pollak1966survival}, showed (again in one dimension) 
that this critical patch size also holds in a stochastic framework.
\citet{barton1987establishment} also showed that
for patches above this critical size,
the probability of establishment of a new mutant that appears at distance $r$ from the patch
decays exponentially with distance from the patch, with the scale given by $\sigma/\sqrt{s_m}$,
and that mutations appearing within the patch have probability of establishment
less than the probability for a panimictic population,
but approaching this as the size of the patch increases.
(This latter panmictic probability of establishment we denote $p_e$,
and often approximate by $2 s_p / \xi^2$ \citep{haldane,fisher}.)
This result holds quite generally for a geographically spread population that experiences a uniform selection
pressure \citep{maruyama1970fixation,cherry2003diffusion}.

@@

\begin{figure}[ht!]
  \begin{center}
    \includegraphics{sim-snapshots}
  \end{center}
  \caption{
  Results from individual-based simulation, on a one-dimensional grid:
  the focal allele is beneficial in the central 20 demes, with $s_b=0.1$,
  and deleterious elsewhere, with $s_m=-0.02$.
  The dynamics were simulated for 10000 generations; demes had 1000 individuals in each;
  dispersal was to nearby demes with a mean dispersal distance of $\sigma \approx 1.66$  times the deme spacing.
  \textbf{(A)} Allele frequencies across space: colored lines are snapshots at an evenly spaced set of times,
  and the solid lines shows the cumulative mean occupation frequency across all generations.
  \textbf{(B)} A segment of the temporal dynamics: space is again shown across the horizontal axis;
  with time on the vertical axis; demes are colored darker red the greater then allele frequency at the corresponding time.
  The 50\% and 5\% frequency contour lines are in grey,
  and the genealogy of 50 individuals in a rare long-distance excursion is traced back with black lines.
  \label{fig:sim_snapshots}
  }
\end{figure}

@@

The arguments depends on the following decomposition of the genealogy of $B$ alleles.
First, fix some intermediate distance $r_0$ from the established patch,
and classify all $B$ alleles present at any time outside $r_0$ into families
if they share a common ancestor living outside $r_0$.
If $r_0$ is large enough that there are not too many families
and interactions between family members are weak,
then these families can be approximated by a collection of independent spatial branching processes.
(This can be made formal in the limit of large population density.)
We can extend these families by 
including any individual outside the original patch (but perhaps inside $r_0$) 
that has a parent outside $r_0$,
so that these branching processes are rooted at $r_0$, and are killed upon reaching the original patch.
This decomposition is illustrated in~\ref{fig:sim_snapshots}, and later in figure~\ref{fig:branching_decomp}.
In terms of these families, the equilibrium frequency above is then
\begin{align}
    \label{eqn:gestalt_q}
    \begin{split}
        q(r) &= \text{ ( outflux of families ) } \times \text{ ( mean occupation density of a family at $r$ ) } ,
\end{split}
\end{align}
and the quantity we wish to compute,
the effective rate of at which \emph{families} of migrant $B$ alleles establish in a new patch at distance $r$, 
is
\begin{align}
    \label{eqn:gestalt_migrate}
    \begin{split}
        \migrate(r) &= \text{ ( outflux of families ) } \times \text{ ( probability a family establishes in patch at $r$ ) }
    \end{split}
\end{align}
All terms except the ``outflux of families'' are simple calculations with spatial branching processes;
and we can solve for this outflux using equation \eqref{eqn:eqfreq}.
Here $\migrate$ reflects the mean number of establishing migrant families per unit time,
rather than individuals,
because family establishment events may include more than one individual,
so individual establishment events are clustered in time,
and using their overall density to obtain a typical waiting time until the next event
will underestimate that time.

@@

To understand the process of adaptation by migration,
we therefore need to look more closely at the rare families of $B$ the allele far from the patches where it is established.
Consider one such family.
As noted above, we can model this by assuming that each individual
migrates, then reproduces, independently of all other family members.
The mean squared migration distance is $\sigma^2$, takes the migrant in a random direction,
and if migration takes the individual into a patch where the $B$ allele is favored, 
we drop the individual from our family.
Since the $B$ allele is uniformly deleterious elsewhere,
we can decouple reproduction and migration
by first forming a tree in which each individual has a random number of offspring
(with mean $1-s_m$ and variance $\xi^2$),
then assigning spatial locations to each individual in the tree,
and then pruning any branches that happen to wander into a patch.
This pruning removes both 
branches leading into the new patch (we will need to count these),
and branches leading back into the already established patch 
(which contribute little, since they are going the wrong direction;
so we ignore these).  % but could add detail in appendix

Since each individual has on average $(1-s_m)$ offspring,
a family has on average $(1-s_m)^t$ members in generation $t$.
Let $p_e(t)$ be the chance of extinction by time $t$, and let $K_t$ be the (random) family size conditioned on nonextinction,
so that $(1-s_m)^t = (1-p_e(t))\E[K_t]$.
It turns out that by assuming the fate of each offspring is independent (i.e.\ that this forms a branching process),
and that the distribution of offspring numbers is not too heavy-tailed \citep[see][for details]{jagers1975branching},
that $K_t$ has a limiting distribution: $K_t \dconv K$ 
and that $(1-s_m)^{t}/(1-p_e(t)) \to \E[K]$ as $t \to \infty$.
In other words, the chance of survival to time $t$ is a constant multiplied by $(1-s_m)^t$,
and conditional on survival, the family size is of fixed magnitude.

We can understand this quite concretely, by a method described rigorously by \citet{geiger1999elementary}.
When we condition on survival until $t$, we condition on existence of at least one lineage from time $0$ to time $t$.
Once given this lineage, nothing distinguishes any of the other reproduction events --
each individual in this lineage must give birth to at least one offspring,
and all other reproduction events are unconditioned.
In other words, the genealogy of a family that survives until $t$
can be decomposed into a ``trunk'' that 
sprouts independent ``branches'' that may or may not survive until $t$.
This process is depicted in Figure~\ref{fig:branching_decomp}A.
Concretely, if we write $Z_t$ for the number of offspring in the branching process alive at time $t$
(so $\P\{Z_t = 0\} = p_e(t)$),
then
\begin{align}
  \left( Z_t \; \vert \; Z_t>0 \right) \deq \sum_{s=0}^t \sum_{k=1}^{W_s-1} Z^{(s,k)}_{t-s},
\end{align}
where each $Z^{(s,k)}$ is an independent copy of $Z$,
and each $W_s$ is an independent copy of the offspring distribution, conditioned on $W_s \ge 1$.

@@

This suggests that we will lose little by following only the motion of the ``trunk''
(i.e.\ the red line in figure \ref{fig:branching_decomp}),
and adding the other $K-1$ surviving family members as a cluster about this trunk.
The motion is is approximately Brownian, with variance $\sigma^2$ per generation,
with $\sigma$ is the dispersal distance.
Suppose that the new patch is at distance $r$.
The probability density for the first time such a Brownian motion $B_s$ hits $r$ is
\citep[XXX]{feller}
\plr{work in $d=2$}
% note $\sigma B_t \deq B_{\sigma^2 t}$ so $\P\{ \sup_{s\le t} \sigma B_s \le r \} = \P\{ \sup_{s \le \sigma^2 t} B_s \le r \}
\begin{align} \label{eqn:hitting_dist}
  \P\{ B_t=r \;\mbox{and} \; B_u<r \;\mbox{for all}\; u<t\} =  \frac{r}{\sigma^3 t^{3/2}\sqrt{2\pi}} \exp\left(-\frac{r^2}{2t\sigma^2}\right) .
\end{align}
Let $\tau$ be this random time at which a given migrant family hits the new patch,
with the convention that $\tau =\infty$ if the family never gets here.
We approximate the chance that a given family arrives at the new patch at time $t$, i.e.\ that $\tau=t$,
by the product of equation~\eqref{eqn:hitting_dist} and the probability that the migrant family survives until time $t$.
For convenience, define $\alpha := - \log(1-s_m)$ to be the rate of decrease between patches (note that $\alpha \approx s_m$),
and $\mu_K = \E[K]$ to be the limiting mean family size conditioned on survival,
so that $1-p_e(t) \simeq (1/\mu_K) \exp(-\alpha t)$.
Then, the probability distribution of $\tau$ is given by
\begin{align}
    \P\{ \tau \in dt \} =  \frac{1}{\mu_K} \frac{r}{\sigma^3 t^{3/2}\sqrt{2\pi}} \exp\left(-\frac{r^2}{2t\sigma^2} - \alpha t \right) dt  .
\end{align}

@@

In summary, the probability that an outmigrant from the already-established patch establishes in a new patch at distance $r$ away is 
equal to the probably the trunk arrives (equation~\eqref{eqn:estab_prob}) multiplied by the probability it establishes (equation~\eqref{eqn:prob_estab_arrival}),
namely
\begin{align} \label{eqn:bp_prob_estab}
    F(r) &:= \text{( prob family establishes at $r$ )} = \frac{2s_p}{\xi^2} \exp\left( - \frac{r \sqrt{2 \alpha}}{\sigma} \right) ,
\end{align}

@@

Now, we can obtain the effective ``outflux of families'' from information about the migrant families, combined with \eqref{eqn:gestalt_q}.
Define $\gamma$ to be this outflux; 
i.e.\ the mean number of individuals per unit time whose ancestors carrying the $B$ allele all lived closer than $r_0$ to the original patch
but migrate to a larger distance.
Then for $r>r_0$, the mean occupation density $q(r)$ is equal to $\gamma$ 
multiplied by the mean occupation density of the family descending from such a migrant,
which we denote $u(r)$.
Denote by $Z_t$ the number of individuals such a family has after time $t$,
and the locations of these by $X_1(t), \ldots, X_{Z_t}(t)$.
Then this mean occupation density, at $x$, is the expected number of individuals ever present at $x$. 
Since the expected number of individuals present at time $t$ is $e^{\alpha t}$
and the location of each of these individuals is marginally distributed as a  Gaussian with variance $\sigma^2 t$,
we can write this expectation as
\begin{align}
    u(x) &= \int_0^\infty \E[ \#\{ k : \; X_k(t) \in dx \} ] dt \\
    &= \int_0^\infty e^{\alpha t} \frac{1}{\sqrt{2 \pi t \sigma^2}} e^{-\frac{x^2}{2t\sigma^2}} dt \\
    &= \frac{ 1 }{ \sigma \sqrt{2 \alpha} } \exp\left( - \frac{ x \sqrt{2\alpha} }{ \sigma } \right) 
\end{align}
\plr{check that}
The integral is evaluated in Appendix~\ref{apx:some_integrals}.
\plr{Also examine non-dependence on $r_0$.}
